# Reproducible Research: Peer Assessment 1

```{r }
library(ggplot2)
```

## Loading and preprocessing the data

```{r}
filePath <- paste0(getwd(),"/activity.zip")

datset <- read.csv(unz(filePath,"activity.csv"))
```

## What is mean total number of steps taken per day?

### For this part of the assignment, you can ignore the missing values in the dataset.

1. Make a histogram of the total number of steps taken each day

2. Calculate and report the mean and median total number of steps taken per day



### 1.
```{r}

totalDaySteps <- aggregate(steps ~ date,data = datset, FUN =sum, na.rm=FALSE)

hist(totalDaySteps$steps, col ="green", xlab="Steps",ylab= "Frequency",
     main="Total Number Of Steps Taken Each Day")

```


### 2.
```{r}

myMean <- mean(totalDaySteps$steps,na.rm=TRUE)
myMedian <-median(totalDaySteps$steps,na.rm=TRUE)
```

The mean for total number of steps taken per day is `r myMean` .

The median for total number of steps taken per day is `r myMedian` .

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


### 1.
```{r}
avgIntervalSteps <- aggregate(steps ~ interval, data= datset, FUN=mean, na.rm=TRUE)


plot <- ggplot(data=avgIntervalSteps,aes(x=interval,y=steps)) + 
        geom_line(col="red",lwd=0.9) +
        scale_x_continuous(breaks = seq(0,2355,150)) + 
        scale_y_continuous(breaks = seq(0,206,50)) +
        theme(axis.text.x=element_text(size=10,colour="blue",
                                       face="bold")) +
        theme(axis.text.y=element_text(colour="blue",
                                       face="bold")) +
        ggtitle("The Average Number Of Steps By 5 Minute Intervals Across All Days .") +
        theme(plot.title=element_text(size = 12,face="bold"))
print(plot)   

```

### 2.
```{r}
maximus <- avgIntervalSteps[which(avgIntervalSteps$steps == max(avgIntervalSteps$steps)),] 
maximus

```
The `r maximus[1]`th interval contains the maximum number of steps. 

## Imputing missing values

**NOTE: There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.**

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
Do these values differ from the estimates from the first part of the assignment? 
What is the impact of imputing missing data on the estimates of the total daily number of steps?

### 1.
```{r}
totalNA <- table(complete.cases(datset))
names(totalNA) <- c("NArows","nonNArows")
totalNA
```
The total number of rows with NAs is `r totalNA[1]`.

### 2.

**The strategy devised for filling in all of the missing values in the dataset consists on identifying the dates (YYYY-MM-DD) when/where steps data is Not Available (Assuming steps as the only column with missing values coded as NA).**

**Dates are reformatted to YYYY-DD in order to compare and fill in its step values with the value from same day(s) (%d) but from different month(s) during that same year that my not have missing values.** 

**Example, if date "2012-11-30" contains missing value, then it may be filled with the steps value from the same day(s) but 
from a different month(s) of the year, in this case it can be filled in from the "2012-10-30" day. An identified day with 
NA can also be filled in by calculating the mean from more than one month that contains the same day. If all same days 
(for all months) contain NA or no monthly equivalent day(s) are found, steps is assigned zero. Below stategy is implemented
and dataset with the missing data filled in (datset2) is created.**


### 3.
```{r}
datset2 <- datset 
tempset <- datset2 
## Change date format to "YYYY-DD" in temporary data set.
tempset$date <- format(as.Date(tempset$date),"%Y-%d")
## ID rows with missing values in temporary data set.
naRow <- !complete.cases(tempset)
## ID dates that contain missing values in temporary data set.
naDates <- unique(tempset[naRow,"date"])

for (i in naDates) {
        ## ID indexes of all rows for a specific date(s) of format "YYYY-DD" in temporary data set.          
        idx <- grep(i,tempset[,2])
        
        ## If all same days (but from different months) contain NA or no monthly 
        ## equivalent day(s) are found, steps value is assigned zero.
        if (all(is.na(tempset[idx,1]))) {
                
                
                datset2[idx,1] <- 0
                
                } else {
                        
                        ## Else calculate steps value from mean of other months (but same day) and assign it to steps columns with NA.       
                        myMean <- mean(as.vector(tempset[idx,1]),na.rm=TRUE)
                        idx2 <- as.integer(which(is.na(tempset[idx,1]), arr.ind=TRUE))
                        datset2[idx[idx2],1] <- myMean
                        
                        }
        
        
        }
```

### 4.
```{r}
totalDaySteps <- aggregate(steps ~ date,data = datset2, FUN =sum, na.rm=FALSE)

hist(totalDaySteps$steps, col ="blue", xlab="Steps",ylab= "Frequency",
     main="Total Number Of Steps Taken Each Day (Filled)")

myMean <- mean(totalDaySteps$steps,na.rm=TRUE)
myMedian <-median(totalDaySteps$steps,na.rm=TRUE)


meTbl <- cbind(myMean,myMedian)
kable(meTbl,format="markdown")
```
**Yes, these last mean/median calculations, for total number of steps taken per day, differ from the estimates from the 1st part of the assignment. This may be in part because of the inclusion/filling strategy used on NA values. Even the histograms differ,
with bars of last histogram having increased in frequency.**

**In this case, the impact of imputing missing values seems minimal, but is more apparent on the mean than the median. In this part by including/filling in the NA values with estimates the mean/median values decreased, but this is more apparent on the
mean value. We were still able to include step values for all dates under the two month period.** 

## Are there differences in activity patterns between weekdays and weekends?

### For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

### 1.
```{r}

datset2$weekdays <- as.factor(ifelse(weekdays(as.Date(datset2$date)) 
                                     %in% c("Saturday","Sunday"),"Weekend","Weekday"))

```

### 2.
```{r }
myMean <- with(datset2,aggregate(steps,by=list(interval,weekdays),FUN = mean, na.rm=FALSE))
colnames(myMean) <- c("interval","weekdays","steps")

plot <-ggplot(data=myMean, aes(x =interval,y =steps)) +
        geom_line(col="red",lwd=0.9) +
        facet_wrap( ~ weekdays, nrow=2,ncol=1) +
        scale_x_continuous(breaks = seq(0,2355,150)) + 
        scale_y_continuous(breaks = seq(0,205,50)) +
        theme(axis.text.x=element_text(size=10,colour="blue",
                                       angle=45,face="bold")) +
        theme(axis.text.y=element_text(colour="blue",
                                       face="bold")) +
        ggtitle("The Average Number Of Steps By 5-Minute Intervals During Weekdays/Weekends.") +
        theme(plot.title=element_text(size = 12,face="bold"))
print(plot)              

```
